Building Trust in Science Through a Bayesian Perspective on Probability and Uncertainty

### 2021-04-29 manuscript.tex
update

\EJ{I have doubts about the ``Trust me, I'm a Bayesian'' part of the title, also because ``Trust'' now occurs twice. Maybe simply ``Building Trust in Science Through a Bayesian Perspective on Probability and Uncertainty''?}
\MK{Ok with the new title - I liked the sound of the trust me thing but utlimatley it may not be really on point!}
\EJ{I also uploaded my bibtex file, ``referenties.bib''. I could not get it to work, but you could copy the relevant references into your own bibtex file. Actually, I also managed to make the initial bibtex setup not work, and I have absolutely no clue as to how this happened.}

@@

\EJ{Yeah so as a hard-nosed determinist I still have an issue with ``inherent'' uncertainty. If we don't sort this out we could resolve it through a footnote (e.g., ``EJ Wagenmakers would like to note that he is a determinist who believes that the concepts of probability and uncertainty belongs wholly to the mind \parencite{Jevons18741913}, with the possible exception of the strange world of quantum mechanics.''} 
\MK{I got rid of the inherent part.}

\EJ{It would be good to go over the reference list in detail; in particular, I noticed several ``bayes'' that should be ``Bayes'' (the B needs to be enclosed in brackets so it displays in upper case)}
Negotiating these many forms of uncertainty in argumentation—and building toward consensus at the same time that uncertainty is present—is a key part of the scientific process \parencite{t00}. In consequence, some philosophers of science have suggested that science is a process that builds increasingly better models which allows us to make increasingly more accurate predictions about the world  \parencite{g10, g06, n02, r77}.

@@

The remainder of this paper is structured in the following way. We will first briefly introduce Bayes theorem and wider ideas around Bayesian epistemology. Next, we will elaborate on the importance of probability and uncertainty in science (education) and review how people learn about these concepts and where they face difficulties. Then, we will provide examples of how Bayes could help to address these issues before concluding with implications and directions for future research.

@@

To make this more tangible, consider the following example. Suppose you are interested in the proportion of students that use Windows vs. MacOS as their operating system. The proportion of Windows users $\theta$ ranges between 0 and 1. Before you start our Bayesian analysis, you need to quantify your background knowledge -- the relative plausibility of the different values for $\theta$. Given your previous experiences in class rooms your best guess is that about half of the students use MacOS. You would not be surprised if the  proportion of Windows users was between 30\% and 70\% but you would be rather surprised to find either no Windows or exclusive Windows usage. This background knowledge motivates the use of the dome-shaped prior distribution shown as the dashed line in Figure~\ref{fig:WindowsMacPriorPosterior}.\footnote{This is the beta$(2,2)$ distribution.} Now suppose you start to collect data by asking students about their operating systems. The first ten observations of Windows (W) and MacOS (M) usage that you make are $\{W, W, M, W, M, W, W, M, W, M\}$: six students use Windows, four use MacOS. With these data in hand, we then combine our prior knowledge (i.e., the dome-shaped prior distribution) with the predictive updating factor (assuming a binomial likelihood) and obtain the posterior distribution shown as the solid line in Figure~\ref{fig:WindowsMacPriorPosterior}. The posterior distribution is more peaked than the prior distribution, indicating that the data have sharpened our knowledge about $\theta$. In addition, the posterior distribution is higher than the prior distribution for values of $\theta$ between approximately $.4$ and $.75$: these values of $\theta$ predicted the data relatively well and have therefore gained credibility. In contrast, values of $\theta$ lower than $.4$ and higher than $.75$ predicted the data relatively poorly, which is why they have lost credibility compared to the prior.\footnote{Note that the area under a probability distribution equals 1, so that posterior gains for specific regions of $\theta$ cancel out exactly against posterior losses for the other regions.}

@@

But how do the principles from this example translate to science (education)? A recent case from physics can serve to illustrate this. In 2011, physicists reported that they had observed elementary particles moving faster than the speed of light \parencite{b11} - and thus defying Einstein’s theory of relativity. However, instead of celebrating a historic discovery, the scientists started to look for what had gone wrong in their observations.

@@

\JR{may cut these depending on what E.J. adds below}
\EJ{OK I added a bunch of stuff. Feel free to ignore, cut, rewrite, integrate with other materials etc. Here is where my stuff starts...}

The famous French astronomer and mathematician Pierre-Simon Laplace argued that Bayes' rule quantifies how rational observers ought to change their opinion in light of new evidence; in this sense, Laplace believed that Bayes' rule was \emph{common sense expressed in numbers}. We now showcase several insights that follow directly from Bayes' rule. In our earlier exposition of Bayes' rule we already emphasized the idea that hypotheses gain credibility when they predict the data well, and lose credibility when they predict new data poorly \parencite{WagenmakersEtAl2016CD}. This idea generalizes what the mathematician George Polya termed the ``Fundamental Inductive Pattern'':
\begin{quotation}
``This inductive pattern says nothing surprising. On the contrary, it expresses a belief which no reasonable person seems to doubt: \emph{The verification of a consequence renders a conjecture more credible}. With a little attention, we can observe countless reasonings in everyday life, in the law courts, in science, etc., which appear to confirm to our pattern.'' \parencite[pp. 4-5]{Polya1954Vol2}
\end{quotation}

@@

Finally, we can also write Bayes' rule as follows:
\begin{equation}
\begin{split}
    p(\mathcal{H}_1 \mid \text{data})& = \frac{p(\text{data} \mid \mathcal{H}_1) p(\mathcal{H}_1)}{p(\text{data})}\\
    & = \frac{p(\text{data} \mid \mathcal{H}_1) p(\mathcal{H}_1)}{p(\text{data} \mid \mathcal{H}_1) p(\mathcal{H}_1) + p(\text{data} \mid \mathcal{H}_0) p(\mathcal{H}_0)}.
\end{split}
\end{equation}
This equation shows that the posterior plausibility is dictated by the predictive performance for $\mathcal{H}_1$ and $\mathcal{H}_0$, weighted by the prior plausibility of each hypothesis. In other words, ``The Bayesian world is a comparative world in which there are no absolutes.'' \parencite[p. 308]{Lindley2000}. This is in stark contrast to $p$-value statistical hypothesis testing, in which a statistical model (the null hypothesis) is judged in isolation. 
\EJ{...and here it ends.}

First, the result is a non-binary inference, one that doesn't speak to whether or not the vaccine is safe, but to what degree it is safe, doing so with a language of probability and uncertainty. Though a small difference, discussing the safety of vaccines on these terms can provide a more solid foundation for principled disagreements and arguments.

Second, Bayes provides a heuristic and a language for engaging with uncertainty in a science literacy context in a principled way that is in-line with how most of us likely already interpret data. Specifically Bayes provides people with a way to talk about weighing between ideas and their prior beliefs and new information and data. In short, Bayes gives people a chance to deal with uncertainty in a principled but accessible way.

Last, his process could be formalized mathematically.

%I think there are several very general ideas that could be illustrated with Bayes’ rule. These include:

%\begin{enumerate}
%    \item When we gain new information, our updated knowledge of the world is a compromise between our old knowledge and the new information.
%    \item Today’s posterior is tomorrow’s prior 
%    \item Hypotheses/claims that predict new data well receive a boost in credibility, whereas those that predict the data poorly suffer a decline.
%    \item All knowledge is relative
%    \item Avoid absolute beliefs (Cromwell’s rule; these are not amenable to learning)
%    \item Surprise lost is credibility gained
%    \item Polya’s fundamental inductive pattern: ``This inductive pattern says nothing surprising. On the contrary, it expresses a belief which no reasonable person seems to doubt: The verification of a consequence renders a conjecture more credible. With a little attention, we can observe countless reasonings in everyday life, in the law courts, in science, etc., which appear to confirm to our pattern.'' (Polya, 1954, pp. 4-5)
%    \item Extraordinary claims require extraordinary evidence
%\end{enumerate}

@@

\appendix

\section{code}
add JASP links/code/etc.?

>>>>

@@

Negotiating these many forms of uncertainty in argumentation—and building toward consensus at the same time that uncertainty is present—is a key part of the scientific process \parencite{t00}. In consequence, some philosophers of science have suggested that science is a process that builds increasingly better models which allows us to make increasingly more accurate predictions about the world  \parencite{g10, g06, n02, r77}.

@@

The remainder of this paper is structured as follows. First, we will introduce Bayes theorem and wider ideas around Bayesian epistemology. Next, we will elaborate on the importance of probability and uncertainty in science (education) and review how people learn about these concepts and where they face difficulties. Then, we will provide examples of how Bayes could help to address these issues before concluding with implications and directions for future research.

@@

To make this more tangible, consider the following example. Suppose an educator is interested in the proportion of students that use Windows or MacOS as their operating system. The proportion of Windows users $\theta$ ranges between 0 and 1. Before you start our Bayesian analysis, you need to quantify your background knowledge -- the relative plausibility of the different values for $\theta$. Given their previous experiences as a teacher, their best guess is that about half of the students use MacOS. They would not be surprised if the proportion of Windows users was between 30\% and 70\% but you would be rather surprised to find either no Windows or exclusive Windows usage.

This background knowledge motivates the use of the dome-shaped prior distribution shown as the dashed line in Figure~\ref{fig:WindowsMacPriorPosterior}.\footnote{This is the beta$(2,2)$ distribution.} Now suppose you start to collect data by asking students about their operating systems. The first ten observations of Windows (W) and MacOS (M) usage that you make are $\{W, W, M, W, M, W, W, M, W, M\}$: six students use Windows, four use MacOS.

With these data in hand, we then combine our prior knowledge (i.e., the dome-shaped prior distribution) with the predictive updating factor (assuming a binomial likelihood) and obtain the posterior distribution shown as the solid line in Figure~\ref{fig:WindowsMacPriorPosterior}. The posterior distribution is more peaked than the prior distribution, indicating that the data have sharpened our knowledge about $\theta$. In addition, the posterior distribution is higher than the prior distribution for values of $\theta$ between approximately $.4$ and $.75$: these values of $\theta$ predicted the data relatively well and have therefore gained credibility. In contrast, values of $\theta$ lower than $.4$ and higher than $.75$ predicted the data relatively poorly, which is why they have lost credibility compared to the prior.\footnote{Note that the area under a probability distribution equals 1, so that posterior gains for specific regions of $\theta$ cancel out exactly against posterior losses for the other regions.}

@@

How do the principles from this example translate to students' work with data and evidence -- not only educators? A recent case from physics can serve to illustrate this approach in the context of students' activity in science classrooms. In 2011, physicists reported that they had observed elementary particles moving faster than the speed of light \parencite{b11} - and thus defying Einstein’s theory of relativity. However, instead of celebrating a historic discovery, the scientists started to look for what had gone wrong in their observations.

@@

The famous French astronomer and mathematician Laplace argued that Bayes' rule quantifies how rational observers ought to change their opinion in light of new evidence; in this sense, Laplace believed that Bayes' rule was \emph{common sense expressed in numbers}. We now showcase several insights that follow directly from Bayes' rule. In our earlier exposition of Bayes' rule we already emphasized the idea that hypotheses gain credibility when they predict the data well, and lose credibility when they predict new data poorly \parencite{WagenmakersEtAl2016CD}. This idea generalizes what the mathematician Polya termed the ``Fundamental Inductive Pattern'':
\begin{quotation}
``This inductive pattern says nothing surprising. On the contrary, it expresses a belief which no reasonable person seems to doubt: \emph{The verification of a consequence renders a conjecture more credible}. With a little attention, we can observe countless reasonings in everyday life, in the law courts, in science, etc., which appear to confirm to our pattern.'' \parencite[pp. 4-5]{Polya1954Vol2}
\end{quotation}

@@

Finally, we can also write Bayes' rule as follows:
\begin{equation}
\begin{split}
    p(\mathcal{H}_1 \mid \text{data})& = \frac{p(\text{data} \mid \mathcal{H}_1) p(\mathcal{H}_1)}{p(\text{data})}\\
    & = \frac{p(\text{data} \mid \mathcal{H}_1) p(\mathcal{H}_1)}{p(\text{data} \mid \mathcal{H}_1) p(\mathcal{H}_1) + p(\text{data} \mid \mathcal{H}_0) p(\mathcal{H}_0)}.
\end{split}
\end{equation}
This equation shows that the posterior plausibility is dictated by the predictive performance for $\mathcal{H}_1$ and $\mathcal{H}_0$, weighted by the prior plausibility of each hypothesis. In other words, ``The Bayesian world is a comparative world in which there are no absolutes.'' \parencite[p. 308]{Lindley2000}. This is in stark contrast to $p$-value statistical hypothesis testing, in which a statistical model (the null hypothesis) is judged in isolation.

Thus, in summary, the result of applying Bayes rule is a non-binary inference, one that doesn't speak to whether or not the vaccine is safe, but to what degree it is safe, doing so with a language of probability and uncertainty. Though a small difference, discussing the safety of vaccines on these terms can provide a more solid foundation for principled disagreements and arguments. In addition, Bayes provides a heuristic and a language for engaging with uncertainty in a science literacy context in a principled way that is in-line with how most of us likely already interpret data. Specifically Bayes provides people with a way to talk about weighing between ideas and their prior beliefs and new information and data. In short, Bayes gives people a chance to deal with uncertainty in a principled but accessible way.

@@
